Finiteness in Cubical Type Theory

### 2020-06-18 sections/finiteness-predicates.tex
more agda instead of mathematical notation

We used some extra types in the above definition, which we will define here:

\begin{definition}[\(\mathbf{List}\)] \label{List}
  In this paper we define lists as \emph{containers}:
  \begin{agdalisting}
    \ExecuteMetaData[agda/Container/List.tex]{list-def}
  \end{agdalisting}
  This may seem quite foreign or complex in comparison to the usual definition
  of lists as an inductive data type:
  \begin{agdalisting}
    \ExecuteMetaData[agda/Data/List/Base.tex]{list-def}
  \end{agdalisting}
  But it turns out to be quite useful for later proofs.

  Moreover, the inductive type and the container-based type are equivalent, as
  is proven in our formalisation.
\end{definition}
\begin{definition}[Containers] \label{container-def}
  A container \cite{abbottContainersConstructingStrictly2005} is a pair
  \(S , P\) where \(S\) is a type, the elements of which are called
  the \emph{shapes} of the container, and \(P\) is a type family on \(S\), where
  the elements of \(P(s)\) are called the \emph{positions} of a container.
  We ``interpret'' a container into a functor defined like so:
  \begin{equation} \label{container-interp}
    \llbracket S , P \rrbracket(A) \coloneqq \Sigma {(s : S)} , \left( P(s) \rightarrow A \right)
  \end{equation}
  Membership of a container can be defined like so:
  \begin{equation} \label{container-membership}
    x \in \mathit{xs} \coloneqq \text{fib}_{\text{snd}(\mathit{xs})}(x)
  \end{equation}
\end{definition}

\begin{definition}[Disjoint Union]
  We define disjoint union as an inductive type.

@@

  Although the inductive type definition is slightly more ergonomic.
\end{definition}
\begin{definition}[\(\mathbf{Fin}\)] \label{Fin}
  \(\mathbf{Fin}(n)\) is the type of natural numbers smaller than \(n\). We
  define it the standard way:
  \begin{equation}
    \begin{alignedat}{3}
      & \mathbf{Fin}(0)     && \coloneqq \bot ; \\
      & \mathbf{Fin}(n + 1) && \coloneqq \top \uplus \mathbf{Fin}(n) ;
    \end{alignedat}
  \end{equation} 
\end{definition}
\subsection{Instances}
Now that we have a suitable definition of finiteness, we will next prove that
some things are finite.
\begin{lemma}
  \(\bot\), \(\top\), and \(\mathbf{Bool}\) are split enumerable.
\end{lemma}
\begin{proof}
  These three types are represented by fairly simple finite sets:
  \begin{align}
    \bot &= \left\{  \right\} \\
    \top &= \left\{ \text{tt}  \right\} \\
    \mathbf{Bool} &= \left\{  \text{false}, \text{true} \right\}
  \end{align}
  Which leads to quite simple proofs of finiteness.

  The support list for is \(\left[  \right]\), \(\left[ \text{tt} \right]\), and
  \(\left[ \text{false}, \text{true} \right]\), respectively, and the cover
  proof is a function that returns the index of the supplied element.
\end{proof}
\begin{agdalisting}
  The code in our formalisation of these three proofs is quite simple.
  Just as an example, the proof that \(\mathbf{Bool}\) is split enumerable is as
  follows:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst}
\end{agdalisting}
With the most basic simple types out of the way, the obvious next choice is the
(non-dependent) sums and products: \(\uplus\) and \(\times\).
Both of these types can be constructed from the \emph{dependent} sum, however,
so that is the type we will prove finite.
From that we can derive a much wider array of finiteness proofs.

@@

There is a much simpler and more elegant solution: we insist that every
\emph{membership proof} must be unique.
This would disallow a definition of \(\mathcal{E}!(\mathbf{Bool})\) with
duplicates, as there are multiple values which inhabit the type \(\text{false}
\in \left[ \text{false}, \text{true}, \text{false} \right]\).
It also allows us to keep most of the split enumerability definition unchanged,
just adding a condition to the returned membership proof in the cover proof.

@@

As we are working in CuTT, a proof of equivalence between two types gives us the
ability to \emph{transport} proofs from one type to the other.
This is extremely powerful, as we will see.
\subsection{Closure Under \(\Pi\)}
The glaring omission from our closure proofs under type formers so far has been
the \(\Pi\) type: we have not proved closure under functions, dependent or
otherwise.
In MLTT, this is of course not provable: since all of the finiteness predicates
we have seen so far imply decidable equality, and since we don't have any kind
of decidable equality on functions in MLTT, we know that we won't be able to
show that any kind of function is finite; even one like \(\mathbf{Bool}
\rightarrow \mathbf{Bool}\).

@@

  We can prove this goal by showing that \(\mathbf{Tuple}(n, U)\) is split
  enumerable: it is made up of finitely many products of points of \(U\), which
  are themselves split enumerable, and \(\top\), which is also split enumerable.
  Lemma~\ref{split-enum-sigma} shows us that the product of finitely many split
  enumerable types is itself split enumerable, proving our goal.
\end{proof}
\section{Cardinal Finiteness} \label{cardinal-finiteness}
While we have removed some of the unnecessary information from our finiteness
predicates, one piece still remains.
\begin{agdalisting}
  The two following proofs are both valid proofs of the finiteness of
  \(\mathbf{Bool}\), and both do not include any duplicates.
  However they still differ:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst} \smallskip
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-rev}
\end{agdalisting}
Each finiteness predicate so far has contained an \emph{ordering} of the
underlying type.
For our purposes, this is too much information: it means that when constructing
the ``category of finite sets'' later on, instead of each type having one
canonical representative, it will have \(n!\), where \(n\) is the cardinality of
the type\footnotemark.

@@

Next, notice that the support lists of any two proofs of manifest Bishop
finiteness must be permutations of each other.
This will allow us to sort the support list of a proof of cardinal finiteness in
a coherently constant (definition~\ref{prop-trunc},
eliminator~\ref{elim-prop-coh}) way, pulling the support list out from the
truncation.
The cover proof emerges naturally from the definition of the permutation.
\subsection{Closure}
Since we don't have a function of type \(\mathcal{C}(A) \rightarrow
\mathcal{B}(A)\), closure proofs on \(\mathcal{B}\) do not transfer over to
\(\mathcal{C}\) trivially (unlike with \(\mathcal{E}!\) and \(\mathcal{B}\)).
The cases for \(\bot\), \(\top\), and \(\mathbf{Bool}\) are simple to adapt: we
can just propositionally truncate their Bishop finiteness proof.

>>>>

We have used two types there which we have not yet defined:
\(\AgdaDatatype{List}\) and \(\AgdaDatatype{\ensuremath{\in}}\).
We will define them here.

\begin{definition}[\(\mathbf{List}\)] \label{List}
  In this paper we will work with two equivalent definitions of lists.
  The first is the standard definition as an inductive type:
  \begin{agdalisting}
    \ExecuteMetaData[agda/Data/List/Base.tex]{list-def}
  \end{agdalisting}

  The second way to define lists is to define them as a \emph{container}:
  \begin{agdalisting}
    \ExecuteMetaData[agda/Container/List.tex]{list-def}
  \end{agdalisting}
  The reason we use this second strange definition is that it turns out to be
  quite useful in some later proofs.
  We have proven the two types equivalent in our formalisation, however, so we
  can switch between them freely without loss of generality.
\end{definition}

In defining lists we have introduced another concept which needs defining:
\(\AgdaDatatype{Fin}\).
\begin{definition}[\(\mathbf{Fin}\)] \label{Fin}
  \(\mathbf{Fin}(n)\) is the type of natural numbers smaller than \(n\). We
  define it the standard way:
  \begin{agdalisting}
    \ExecuteMetaData[agda/Data/Fin/Base.tex]{fin-def}
  \end{agdalisting}
\end{definition}
Here \(\uplus\) refers to the disjoint union of two types.
\begin{definition}[Disjoint Union]
  We define disjoint union as an inductive type.

@@

  Although the inductive type definition is slightly more ergonomic.
\end{definition}

After that interlude, we can get back to defining containers.
\begin{definition}[Containers] \label{container-def}
  A container \cite{abbottContainersConstructingStrictly2005} is a pair
  \(S , P\) where \(S\) is a type, the elements of which are called
  the \emph{shapes} of the container, and \(P\) is a type family on \(S\), where
  the elements of \(P(s)\) are called the \emph{positions} of a container.
  We ``interpret'' a container into a functor defined like so:
  \begin{agdalisting} \label{container-interp}
    \ExecuteMetaData[agda/Container.tex]{container-interp}
  \end{agdalisting}
  Membership of a container can be defined like so:
  \begin{agdalisting} \label{container-membership}
    \ExecuteMetaData[agda/Container/Membership.tex]{membership-def}
  \end{agdalisting}
\end{definition}

\subsection{Instances}
Now that we have a suitable definition of finiteness, we will next prove that
some things are finite.
\begin{lemma}
  \(\bot\), \(\top\), and \(\AgdaDatatype{Bool}\) are split enumerable.
\end{lemma}
\begin{proof}
  These three types are quite obviously finite: we will show only the proof of
  finiteness for \(\AgdaDatatype{Bool}\) here for brevity's sake.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst}
  \end{agdalisting}
\end{proof}
With the most basic simple types out of the way, the obvious next choice is the
(non-dependent) sums and products: \(\uplus\) and \(\times\).
Both of these types can be constructed from the \emph{dependent} sum, however,
so that is the type we will prove finite.
From that we can derive a much wider array of finiteness proofs.

@@

There is a much simpler and more elegant solution: we insist that every
\emph{membership proof} must be unique.
This would disallow a definition of \(\mathcal{E}!(\AgdaDatatype{Bool})\) with
duplicates, as there are multiple values which inhabit the type \(\text{false}
\in \left[ \text{false}, \text{true}, \text{false} \right]\).
It also allows us to keep most of the split enumerability definition unchanged,
just adding a condition to the returned membership proof in the cover proof.

@@

As we are working in CuTT, a proof of equivalence between two types gives us the
ability to \emph{transport} proofs from one type to the other.
This is extremely powerful, as we will see.
\subsection{Closure Under \(\Pi\)}
The glaring omission from our closure proofs under type formers so far has been
the \(\Pi\) type: we have not proved closure under functions, dependent or
otherwise.
In MLTT, this is of course not provable: since all of the finiteness predicates
we have seen so far imply decidable equality, and since we don't have any kind
of decidable equality on functions in MLTT, we know that we won't be able to
show that any kind of function is finite; even one like \(\AgdaDatatype{Bool}
\rightarrow \AgdaDatatype{Bool}\).

@@

  We can prove this goal by showing that \(\mathbf{Tuple}(n, U)\) is split
  enumerable: it is made up of finitely many products of points of \(U\), which
  are themselves split enumerable, and \(\top\), which is also split enumerable.
  Lemma~\ref{split-enum-sigma} shows us that the product of finitely many split
  enumerable types is itself split enumerable, proving our goal.
\end{proof}
\section{Cardinal Finiteness} \label{cardinal-finiteness}
While we have removed some of the unnecessary information from our finiteness
predicates, one piece still remains.
\begin{agdalisting}
  The two following proofs are both valid proofs of the finiteness of
  \(\AgdaDatatype{Bool}\), and both do not include any duplicates.
  However they still differ:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst} \smallskip
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-rev}
\end{agdalisting}
Each finiteness predicate so far has contained an \emph{ordering} of the
underlying type.
For our purposes, this is too much information: it means that when constructing
the ``category of finite sets'' later on, instead of each type having one
canonical representative, it will have \(n!\), where \(n\) is the cardinality of
the type\footnotemark.

@@

Next, notice that the support lists of any two proofs of manifest Bishop
finiteness must be permutations of each other.
This will allow us to sort the support list of a proof of cardinal finiteness in
a coherently constant (definition~\ref{prop-trunc},
eliminator~\ref{elim-prop-coh}) way, pulling the support list out from the
truncation.
The cover proof emerges naturally from the definition of the permutation.
\subsection{Closure}
Since we don't have a function of type \(\mathcal{C}(A) \rightarrow
\mathcal{B}(A)\), closure proofs on \(\mathcal{B}\) do not transfer over to
\(\mathcal{C}\) trivially (unlike with \(\mathcal{E}!\) and \(\mathcal{B}\)).
The cases for \(\bot\), \(\top\), and \(\AgdaDatatype{Bool}\) are simple to adapt: we
can just propositionally truncate their Bishop finiteness proof.
